---
title: "2- Disaggregation by cause of death"
subtitle: "Bloc A · Level 2 — Epidemiological transitions in Mexico, 2000–2024"
# date: today
# date-format: "MMMM YYYY"
lang: "en"
---

<span class="badge-nivel">Bloque A · Nivel 2</span>
**Variables** : `year` · `sex` · `cause`



## Introduction

Incorporating **cause of death** as an analytical dimension makes it possible to identify the major **epidemiological transitions** characterizing Mexico's health trends over the last three decades.

At this level, the aim is to answer three key questions:

1. Which diseases account for the greatest mortality burden?
2. How has the composition by cause evolved over time?
3. Are there gender differences in cause-specific patterns?

::: {.callout-tip}
**Link to the previous level**  
[Nivel 1 — General description](A1_descriptive_general.qmd) establishes the  mortality trends. Here, we add the causal dimension to explain those trends.
:::


```{r setup-n2, include=FALSE}
 
library(tidyverse)
library(scales)
library(knitr)
library(kableExtra)
library(ggrepel)
options(ggrepel.max.overlaps = Inf)
library(patchwork)   # composition de graphiques
library(dplyr)
library(paletteer)

# ── Données ────────

load("data/asmr_nac_cause.rda")
load("data/deces_cor_imp.rda")
load("data/asmr_cause_corr.rda")


#------------------------- ------------------------------
 cause_labels  <- c(
   "01"="Infectious",
   "02"="Neoplasms",
   "04"="Endocrine",
   "06"="Nervous",
   "09"="Circulatory",
   "10"="Respiratory",
   "11"="Digestive",
   "14"="Genitourinary",
   "16"="Perinatal",
   "18"="Symptoms",
   "20"="External",
   "22"="COVID-19",
   "Other" = "Other causes")
 
 cols <- c(
   "01"="#3E7D44",   # Infectious
   "02"="#8DA0CB",   # Neoplasms
   "04"="#DEA9BB",   # Endocrine
   "06"="#4F6DB8",   # Nervous
   "09"="#CE6A6B",   # Circulatory
   "10"="#2C75B8",   # Respiratory
   "11"="#E5C494",   # Digestive
   "14"="#1B9E77",   # Genitourinary
   "16"="#FDB863",   # Perinatal
   "18"="#A6A6A6",   # Symptoms
   "20"="#8CA893",   # External
   "22"="#A6CEE3",   # COVID
   "Other" = "#CFCFCE")

#------------------------- ------------------------------
deces <- deces_final %>%
   # filter(year >= 2000 & year <= 2024) %>%
   mutate(
     cause = if_else(CAPITULO %in% names(cause_labels)[names(cause_labels) != "Other"],
                     CAPITULO,
                     "Other"),
     cause_label = cause_labels[cause],
       label_sex = recode(sex,
                    "1" = "Men",
                    "2" = "Women")
   )

 
SDR <- asmr_nac_cause |>
   mutate(
     cause = if_else(CAPITULO %in% names(cause_labels)[names(cause_labels) != "Other"],
                     CAPITULO,
                     "Other"),
          cause_label = cause_labels[cause],

       label_sex = recode(sex,
                    "1" = "Men",
                    "2" = "Women")
   )
 
 #------------------------- ------------------------------

mort <- deces |> 
   # filter(year > 1989) |> 
   rename(deaths = n_death)  |>  mutate(year = as.numeric(as.character( year)))

# ── Palette cohérente (définie dans index.qmd, reproduite ici) ───────
palette_sexo <- c("1" = "#2a7a8c", "2" = "#e84855", "Total" = "grey40")


#-------------------------plot SMR ------------------------------

age_levels <- c(
  "0", "1-4", "5-9", "10-14", "15-19", "20-24",
  "25-29", "30-34", "35-39", "40-44", "45-49",
  "50-54", "55-59", "60-64", "65-69", "70-74",
  "75-79", "80-84", "85-89", "90-94", "95-99", "100+"
)
 deces<- deces |> 
mutate(
    age_group = case_when(
      age == 0 ~ "0",
      age >= 1 & age <= 4 ~ "1-4",
      age >= 5 & age <= 9 ~ "5-9",
      age >= 10 & age <= 14 ~ "10-14",
      age >= 15 & age <= 19 ~ "15-19",
      age >= 20 & age <= 24 ~ "20-24",
      age >= 25 & age <= 29 ~ "25-29",
      age >= 30 & age <= 34 ~ "30-34",
      age >= 35 & age <= 39 ~ "35-39",
      age >= 40 & age <= 44 ~ "40-44",
      age >= 45 & age <= 49 ~ "45-49",
      age >= 50 & age <= 54 ~ "50-54",
      age >= 55 & age <= 59 ~ "55-59",
      age >= 60 & age <= 64 ~ "60-64",
      age >= 65 & age <= 69 ~ "65-69",
      age >= 70 & age <= 74 ~ "70-74",
      age >= 75 & age <= 79 ~ "75-79",
      age >= 80 & age <= 84 ~ "80-84",
      age >= 85 & age <= 89 ~ "85-89",
      age >= 90 & age <= 94 ~ "90-94",
      age >= 95 & age <= 99 ~ "95-99",
      age >= 100 ~ "100+"
    ),
    age_group = factor(age_group, levels = age_levels)

  )
#  # Prepare the mapping table for CHAPTER labels
cause_name <- c( "01" = "1- Certain infectious and parasitic diseases",
"02" = "2- Tumors (neoplasms)",
"03" = "3- Diseases of the blood and immune system",
"04" = "4- Endocrine, nutritional, and metabolic diseases",
"05" = "5- Mental and behavioral disorders",
"06" = "6- Diseases of the nervous system",
"07" = "7- Diseases of the eye and adnexa",
"08" = "8- Diseases of the ear and mastoid process",
"09" = "9- Diseases of the circulatory system",
"10" = "10- Diseases of the respiratory system",
"11" = "11- Diseases of the digestive system",
"12" = "12- Skin and subcutaneous tissue",
"13" = "13- Diseases of the musculoskeletal system and connective tissue",
"14" = "14- Diseases of the genitourinary system",
"15" = "15- Diseases of pregnancy, childbirth, and the puerperium",
"16" = "16- Conditions originating in the perinatal period",
"17" = "17- Congenital malformations and chromosomal abnormalities",
"18" = "18- Symptoms, signs, and abnormal clinical and laboratory findings",
"20" = "20- External causes of morbidity and mortality", "22" = "22- COVID-19" )
# ── Données (remplacer par readRDS("data/mort_simulated.rds")) ────────

# ── Agrégations utiles ────────────────────────────────────────────────
# Par année × cause
mort_causa <- mort |>
  group_by(year, cause) |>
  summarise(deaths = sum(deaths), .groups = "drop")

# Par année × cause × sexe
mort_causa_sex <- mort |>
  group_by(year, sex, cause) |>
  summarise(deaths = sum(deaths), .groups = "drop")

# Part relative de chaque cause par an
mort_causa_pct <- mort_causa |>
  group_by(year) |>
  mutate(pct = deaths / sum(deaths) * 100) |>
  ungroup()

# Dernière année disponible
last_year <- max(mort$year)
# class(year)  
```

## Vue d'ensemble : composition de la mortalité

### Évolution de la structure par cause

::: {.panel-tabset}

<!-- #### Aire empilée (%) -->

<!-- ```{r fig-aire-empilada, fig.cap="Structure de la mortalité par cause, 1998–2024 (%)"} -->
<!-- lookup <- data.frame( -->
<!--   cause = as.character(names(cause_name)), -->
<!--   cause_name = as.character(cause_name) -->
<!-- ) -->

<!-- mort_causa_pct <- mort_causa_pct %>% -->
<!--   mutate(cause = as.character(cause)) %>% -->
<!--   left_join(lookup, by = "cause") -->


<!-- ggplot(mort_causa_pct, -->
<!--        aes(x = year, y = pct, fill = cause_name)) + -->
<!--   geom_area(alpha = 0.88, color = "white", linewidth = 0.3) + -->
<!--   scale_fill_manual(values = cols_20) + -->
<!--   scale_x_continuous(breaks = seq(1998, 2022, 4)) + -->
<!--   scale_y_continuous(labels = label_percent(scale = 0.3)) + -->
<!--   labs( -->
<!--     title    = "Composition of mortality by cause group", -->
<!--     subtitle = "Annual proportion of deaths, Mexico 1990–2024", -->
<!--     x = NULL, y = "% deaths", fill = NULL -->
<!--   ) + -->
<!--   theme_minimal(base_size = 13) + -->
<!--   theme( -->
<!--     legend.position  = "bottom", -->
<!--     legend.key.size  = unit(0.2, "cm"), -->
<!--     plot.title       = element_text(face = "bold", color = "#1a2744"), -->
<!--     plot.subtitle    = element_text(color = "#6c757d"), -->
<!--     panel.grid.minor = element_blank() -->
<!--   ) + -->
<!--   guides(fill = guide_legend(nrow = 2)) -->
<!-- ``` -->

#### Absolute trends

```{r fig-lineas-absolutas, fig.cap="Absolute evolution of mortality by cause"}
ggplot(mort_causa,
       aes(x = year, y = deaths / 1000,
           color = cause, group = cause)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2, alpha = 0.8) +
  geom_label_repel(
    data  = mort_causa |> filter(year == last_year),
    aes(label = str_wrap(cause, 18)),
    size  = 3, nudge_x = 0.8, segment.color = "grey60",
    show.legend = FALSE
  ) +
  scale_color_manual(values = cols) +
  scale_x_continuous(breaks = seq(1998,  2024, 4),
                     limits = c(1998, 2025)) +
  scale_y_continuous(labels = label_comma(suffix = "")) +
  labs(
    title    = "Absolute trend in mortality by cause",
subtitle = "Number of deaths (thousands), Mexico 2000–2024",
x = NULL, y = "Deaths (thousands)", color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "none",
    plot.title       = element_text(face = "bold", color = "#1a2744"),
    plot.subtitle    = element_text(color = "#6c757d"),
    panel.grid.minor = element_blank()
  )
```

#### Top causas por sexo  
```{r fig-ev-temp-cause, fig.cap="Temporal evolution of the main causes, 2000–2024"}

# ── 5. Évolution temporelle — top causes ─────────────────────────
top_cap <- deces |>
  group_by(CAPITULO, cause) |>
  summarise(deaths = sum(n_death), .groups = "drop") |>
  slice_max(deaths, n = 10)

top6 <- top_cap |> slice_max(deaths, n = 6) |> pull(CAPITULO)

p6 <- deces |>
  filter(CAPITULO %in% top6) |>
  group_by(year, sex, cause_label) |>
  summarise(deaths = sum(n_death), .groups = "drop") |>
  mutate(cause_name = str_wrap(cause_label, 22)) |>
  ggplot(aes(x = year, y = deaths / 1000,
             color = factor(sex), group = factor(sex))) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ cause_name, scales = "free_y", ncol = 3) +
  scale_color_manual(values = palette_sexo,
                     labels = c("1" = "male", "2" = "female")) +
  scale_x_continuous(breaks = c(2000, 2010, 2020, 2024)) +
  scale_y_continuous(labels = label_comma(suffix = " k")) +
  labs(title = "Trends in the main causes by sex",
       x = NULL, y = "Deaths (thousands)", color = "Sex") +
  theme_minimal(base_size = 10) +
  theme(plot.title     = element_text(face = "bold", color = "#1a2744"),
        legend.position = "bottom",
        strip.text     = element_text(face = "bold", size = 8),
        panel.grid.minor = element_blank())

p6
```

#### Summary

```{r tabla-causa-anio}
mort_causa |>
  filter(year %in% c(2000, 2005, 2010, 2015, 2020, 2024), cause!="Other"
         ) |>
  pivot_wider(names_from = year, values_from = deaths) |>
  arrange(desc(`2024`)) |>
    mutate(cause = unname(cause_name[cause])) |>

  mutate(across(where(is.numeric), ~ format(.x, big.mark = ","))) |>
  kable(
    caption = "Deaths by cause of death, selected years",
    col.names = c("Cause", "2000", "2005", "2010", "2015", "2019", "2024")
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) |>
  row_spec(1, bold = TRUE, background = "#f0f4f8")
  

  
```

:::
## Standardized mortality rates (SMRs)

### SMRs by cause, sex, and year

::: {.panel-tabset}

#### Trends in causes of death (standardized rate)

```{r fig-stdcausedeath, fig.cap="Évolution de la mortalité std par cause"}

# fonctionne
# ggplot(SDR, aes(x = factor(year), y = asmr, fill = cause)) +
#   geom_bar(stat = "identity", position = "stack", width = 0.7) +
#   facet_wrap(~ label_sex) +
#   scale_fill_manual(values = cols[1:length(unique(SDR$cause))], 
#                     labels = cause_labels) +
#   
#   labs(
#     title = "Standardized causes of death (2000-2023)",
#     x = "Years",
#     y = "Standardized rate",
#     fill = "Causes of death"
#   ) +
#   theme_minimal() +
#   theme(
#     axis.text.x      = element_text(angle = 45, hjust = 1, size = 7),
#     axis.text.y      = element_text(size = 7),
#     legend.position  = "bottom",
#     legend.direction = "horizontal",
#   legend.title.position = "top",
#     legend.text      = element_text(size = 6, margin = margin(b = 0)),
#     legend.title     = element_text(size = 7),
#     legend.key.size  = unit(0.3, "cm"),
#     legend.spacing.y = unit(0.05, "cm"),
#     legend.spacing.x = unit(0.05, "cm"),
#     plot.margin      = margin(10, 10, 10, 10)
#   ) +
#   guides(fill = guide_legend(nrow = 2, byrow = TRUE))
 data <- asmr_nac_cause %>%
   # filter(year >= 2000 & year <= 2024) %>%
   mutate(
     cause = if_else(CAPITULO %in% names(cause_labels)[names(cause_labels) != "Other"],
                     CAPITULO,
                     "Other"),
       sex = recode(sex,
                    "1" = "Men",
                    "2" = "Women")
   )
 
 df <- data %>%
   # filter(CAPITULO != "22") %>%
   # filter(year != "2020" & year != "2021" ) %>%
   group_by( year, sex, cause )%>%
   # ANIO, ENT_REGIS, CAPITULO, MORT_STD)
   summarise(asmr = sum(asmr),
             .groups = "drop")
 
 
 # plot
 ggplot(df,
        aes(year,
            asmr,
            fill = cause)) +
   facet_wrap(~sex,
              ncol = 1) +
   geom_bar(stat = "identity", position = "stack") +
   # geom_line(linewidth = 1) +
   facet_wrap(~sex, ncol = 1) +
   scale_fill_manual(
     values = cols,
     labels = cause_labels
   ) +
   labs(
     # title="Main causes of mortality by sex",
     # subtitle="Mexico, 2000–2024",
     x=NULL,
     y="ASMR per 100,000 population",
     fill="Cause"
   ) +
   theme_classic(base_size = 11)+ theme(
     # plot.title=element_text(face="bold", size=18),
     strip.background=element_blank(),
     # strip.text=element_text(face="bold",
     #                         size=14),
     legend.position="right",
     legend.title=element_text(size=12),
     panel.border=element_rect(
       colour="black",
       fill=NA    ),
     axis.text=element_text(colour="black"),
     axis.line=element_line(linewidth=.5)
   )


```
<!-- #### Top 5 des causes de décès (std rate) -->

```{r fig-top5stdcausedeath, fig.cap="Top 5 des causes de la mortalité (std)"}
#--------------------------Identifier les 5 principales causes de décès par année-----------------------------
# top_5_causes_year <- SDR %>%
#   group_by(year, sex) %>%
#   slice_max(SDR, n = 5) %>%
#   ungroup() %>%
#   arrange(year, desc(SDR))
# 
# top_5_causes_year <- top_5_causes_year %>%
#   mutate(
#     capitulo = as.character(CAPITULO),
#     cause_name = libelles_cause[CAPITULO],  # Map codes to names
#     # label_sex = label_sex[sex]
#      sex_label  = case_when(
#       sex == 1 ~ "Men",
#       sex == 2 ~ "Women")
#   )

#-------------------------------------------------------
# 
# ggplot(top_5_causes_year, aes(x = factor(year), y = SDR, fill = cause_name)) +
#   geom_bar(stat = "identity", position = "stack") +
#   facet_wrap(~ sex_label) +
#   # scale_fill_paletteer_d("palettetown::carvanha") +
#   scale_fill_manual(values = cols_7[1:length(unique(SDR$libelles_cause))]) +
#   
#   labs(
#     title = "Top 5 standardized causes of death (2000-2023)",
#     x = "Years",
#     y = "Standardized rate",
#     fill = "Causes of death"
#   ) +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
# 

```

#### Main causes (men)
```{r tabla-causa-anio-asmr-h}


# Étape 1 : Top 5 causes par sexe basé sur 2024
top5_hommes <- SDR |>
  filter(sex == 1, year == 2024) |>
  arrange(desc(asmr)) |>
  slice_head(n = 10) |>
  select(CAPITULO, cause)

# top5_hommes

# Étape 2 : Filtrer les données pour les années sélectionnées
annees <- c(1990, 1995, 2000, 2005, 2010, 2015, 2019, 2024)

data_hommes <- SDR |>
  filter(sex == 1, year %in% annees, CAPITULO %in% top5_hommes$CAPITULO)

# data_hommes

# Étape 3 : Pivot
wide_hommes <- data_hommes |>
  select(cause_label, year, asmr) |>
  pivot_wider(names_from = year, values_from = asmr) |>
  arrange(desc(`2024`))

# afficer tableau hommes !!!
# wide_hommes

# Étape 4 : Tableau
library(knitr)
library(kableExtra)

format_table <- function(wide_data, caption_txt) {
  wide_data |>
    mutate(across(where(is.numeric), ~ format(round(.x, 1), big.mark = ","))) |>
    kable(
      caption   = caption_txt,
      col.names = c("Causes", names(wide_data)[-1]),
      align     = c("l", rep("r", ncol(wide_data) - 1))
    ) |>
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width        = FALSE,
      font_size         = 13
    ) |>
    add_header_above(
      c(" " = 1,
        "Standardized mortality rate (per 100,000 inhab.)" = ncol(wide_data) - 1)
    ) |>
    row_spec(1, bold = TRUE, background = "#f0f4f8")
}

tbl_hommes <- format_table(wide_hommes, "")

tbl_hommes

  
```
#### Main causes (women)
```{r tabla-causa-anio-asmr-m}


# Étape 1 : Top 5 causes par sexe basé sur 2024
top5_femmes <- SDR |>
  filter(sex == 2, year == 2024) |>
  arrange(desc(asmr)) |>
  slice_head(n = 10) |>
  select(CAPITULO, cause_name)


# top5_femmes

# Étape 2 : Filtrer les données pour les années sélectionnées
annees <- c(1990, 1995, 2000, 2005, 2010, 2015, 2019, 2024)

data_femmes <- SDR |>
  filter(sex == 2, year %in% annees, CAPITULO %in% top5_femmes$CAPITULO)

# data_femmes

# Étape 3 : Pivot
wide_femmes <- data_femmes |>
  select(cause_label, year, asmr) |>
  pivot_wider(names_from = year, values_from = asmr) |>
  arrange(desc(`2024`))

# wide_femmes

# Étape 4 : Tableau
library(knitr)
library(kableExtra)

format_table <- function(wide_data, caption_txt) {
  wide_data |>
    mutate(across(where(is.numeric), ~ format(round(.x, 1), big.mark = ","))) |>
    kable(
      caption   = caption_txt,
      col.names = c("Causes", names(wide_data)[-1]),
      align     = c("l", rep("r", ncol(wide_data) - 1))
    ) |>
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width        = FALSE,
      font_size         = 13
    ) |>
    add_header_above(
      c(" " = 1,
        "Standardized mortality rate (per 100,000 inhab.)" = ncol(wide_data) - 1)
    ) |>
    row_spec(1, bold = TRUE, background = "#f0f4f8")
}

tbl_femmes <- format_table(wide_femmes, "")

tbl_femmes
```
:::


<!-- ## Différences de genre dans les patterns causaux -->
  
::: {.panel-tabset}

<!-- #### Hommes vs Femmes (2023) -->

<!-- ```{r fig-sex-causa-barras, fig.cap="Mortalité par cause et sexe, dernière année disponible"} -->
<!-- ``` -->

<!-- #### Ratio H/M par cause -->

<!-- ```{r fig-ratio-hm} -->
<!-- ``` -->

<!-- #### Tendencias por sexo -->

<!-- ```{r fig-tendencias-sexo, fig.height=7, fig.cap="Tendencias por causa y sexo, 2000–2023"} -->
<!-- ``` -->

:::
  

  
## Concentration indicators
  
```{r indice-concentracion}
# Indice de Herfindahl-Hirschman (concentration causale)
ihh <- SDR |>
  group_by(year) |>
  summarise(ihh = sum((n_death / 100)^2),
            n_causas = n_distinct(CAPITULO),
            causa_top = CAPITULO[which.max(n_death)],
            pct_top   = max(n_death)) |>
  ungroup()

ggplot(ihh, aes(x = year, y = ihh)) +
  geom_line(color = "#2a7a8c", linewidth = 1.2) +
  geom_point(color = "#2a7a8c", size = 3) +
  scale_x_continuous(breaks = seq(2000, 2024, 4)) +
  labs(
   title    = "Mortality concentration index by cause (HHI)",
subtitle = "Higher values indicate greater concentration in a few causes",
    x = NULL, y = "Índice Herfindahl-Hirschman"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", color = "#1a2744"),
    plot.subtitle = element_text(color = "#6c757d")
  )
```



## Key points 

::: {.callout-note icon="false"}

- Mortality in Mexico is dominated by **cardiovascular** and **endocrines** diseases, reflecting an advanced epidemiological transition.
- **External causes** (accidents, violence) are overrepresented in male causes.
<!-- show a male-to-female ratio significantly higher than 1, revealing a strong gender component. -->
<!-- - Temporal trends show a gradual **diversificaction** of causes, evident after the decline of the HHI in 2020. -->
:::



## Navigation

::: {style="display:flex; justify-content:space-between; margin-top:2rem;"}
[← 1- General description](A1_descriptive_general.qmd){.btn .btn-outline-secondary}
[→ 3- Specific causes](A3_epecific_causes.qmd){.btn .btn-primary}
:::
