---
title: "Mortality Analysis in Mexico"
subtitle: "A multidimensional approach: temporality, causes and territory"
date: today
# date-format: "MMMM YYYY"
lang: en
sidebar: sidebar-en
---

```{r setup-global-en, include=FALSE, cache=FALSE}
library(tidyverse); library(scales); library(knitr); library(kableExtra)

palette_sex   <- c("1" = "#2a7a8c", "2" = "#e84855", "Total" = "grey40")
# palette_cause <- c(
#   "Cardiovascular diseases"  = "#1a2744",
#   "Malignant tumors"         = "#e84855",
#   "Diabetes mellitus"        = "#2a7a8c",
#   "Respiratory diseases"     = "#f4a261",
#   "External causes"          = "#6c757d",
#   "Other"                    = "#adb5bd"
# )

# ── Real data — update path as needed ───────────────────────────────
# load("data/asmr_nac.rda")
 load("H:/Mon Drive/Broni/Projet R Mortality/BDD/Data_clean/National/20260421 asmr/asmr_nac.rda")

mort <- asmr_nac |> mutate(year = as.integer(year))

# ── Simulated data (demo) ────────────────────────────────────────────
# set.seed(42)
# mort <- expand.grid(year = 1990:2024, sex = c("1", "2")) |>
#   mutate(
#     deaths = rpois(n(), lambda = 400000 + 8000*(year-1990) +
#                      ifelse(sex == "1", 50000, 0) +
#                      ifelse(year %in% 2020:2021, 120000, 0)),
#     asmr   = deaths / runif(n(), 80e6, 130e6) * 1e5
#   )

mort_national <- mort |>
  group_by(year, sex) |>
  summarise(deaths = sum(n_death), .groups = "drop")
```

<!-- Hero banner -->
::: {.column-screen style="background: linear-gradient(135deg, #1a2744 0%, #2a7a8c 100%); padding: 4rem 2rem; margin-bottom: 2.5rem; color: white;"}
::: {.column-body}

## Mortality in Mexico {style="color:white; border-bottom: 2px solid rgba(255,255,255,0.3); padding-bottom:0.5rem;"}

This site presents a systematic and progressive analysis of mortality in Mexico
based on administrative records (INEGI / Ministry of Health),
covering the period **1990–2024**.
The analytical architecture is organised in **three complementary blocks**:

:::{style="display:flex; gap:2rem; margin-top:1.5rem; flex-wrap:wrap;"}

:::{style="flex:1; min-width:240px; background:rgba(255,255,255,0.12); border-radius:8px; padding:1.2rem;"}
**📈 Block A — Epidemiological analysis**  
Progressive disaggregation: global trends → causes →
specific causes → states → municipalities.
:::

:::{style="flex:1; min-width:240px; background:rgba(255,255,255,0.12); border-radius:8px; padding:1.2rem;"}
**🗺️ Block B — Regional disparities**  
Multi-scale mapping: state-level distribution,
municipal hotspots, regional gradients.
:::

:::{style="flex:1; min-width:240px; background:rgba(255,255,255,0.12); border-radius:8px; padding:1.2rem;"}
**⏳ Block C — Life expectancy**  
Trends in life expectancy at birth and at selected ages,
disaggregated by sex, state, and municipality.
:::

:::
:::
:::

## Site structure

### Block A — Health and temporal analysis

| Level | Main variables | Analytical objective | Page |
|:---:|:---|:---|:---:|
| ① | `year`, `sex` | Global trends, ASMR | [→](A1_descriptive_general.qmd) |
| ② | `year`, `sex`, `cause` | Epidemiological transitions | [→](A2_causes.qmd) |
| ③ | `year`, `sex`, `cause_specific` | ICD-10 specific cause breakdown | [→](A3_specific_causes.qmd) |
| ④ | `year`, `sex`, `state`, `cause` | Regional disparities | [→](A4_territorial.qmd) |
| ⑤ | `year`, `sex`, `state`, `cause_specific` | Health-territory articulation | [→](A5_maximum_disaggregation.qmd) |

### Block B — Spatial analysis

| Level | Main variables | Analytical objective | Page |
|:---:|:---|:---|:---:|
| ① | `state`, `cause` | Regional description | [→](B1_regional.qmd) |
| ② | `state`, `year`, `cause` | Spatial distribution, clusters | [→](B2_spatial_states.qmd) |
| ③ | `municipality`, `year`, `cause` | Intra-state heterogeneity | [→](B3_spatial_municipalities.qmd) |

### Block C — Life expectancy

| Level | Main variables | Analytical objective | Page |
|:---:|:---|:---|:---:|
| ① | `year`, `sex` | National overview | [→](C1_life_expectancy.qmd) |
| ② | `state`, `year`, `sex` | State-level trends | [→](C2_le_states.qmd) |
| ③ | `municipality`, `sex` | Municipal heterogeneity | [→](C3_le_municipalities.qmd) |

---

## Preliminary overview

```{r fig-national-overview, fig.cap="Total mortality trends by sex (1990–2024)"}
ggplot(mort_national,
       aes(x = year, y = deaths / 1000, color = sex, group = sex)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_color_manual(values = palette_sex,
                     labels = c("1" = "Men", "2" = "Women")) +
  scale_x_continuous(breaks = seq(1990, 2024, 4)) +
  scale_y_continuous(labels = label_comma(suffix = " k")) +
  labs(
    title    = "Total Mortality in Mexico",
    subtitle = "Number of deaths by sex, 1990–2024",
    x = NULL, y = "Deaths (thousands)", color = "Sex"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", color = "#1a2744"),
    plot.subtitle    = element_text(color = "#6c757d"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

::: {.callout-note}
<!-- **Note on data**   -->
<!-- The values presented here are **simulated** for demonstration purposes. -->
<!-- For the real analysis, replace the `setup-global-en` chunk with: -->
<!-- `load("data/asmr_nac.rda")`. -->
<!-- Recommended sources: INEGI (SINAIS) and the DGE of the SSA. -->
:::

---

## How to navigate this site

1. **Progressive reading** — follow Block A from level ① to ⑤,
   gradually increasing analytical complexity.
2. **Thematic reading** — go directly to Block B (spatial perspective)
   or Block C (life expectancy) depending on your research question.
3. **Language switch** — use the **🇲🇽 Español / 🇬🇧 English** menu
   in the navigation bar to access equivalent content in Spanish.

::: {style="display:flex; justify-content:flex-end; margin-top:2rem;"}
[→ ① General description](A1_descriptive_general.qmd){.btn .btn-primary}
:::
