---
title: "1- Análisis descriptivo general"
subtitle: "Bloque A · Nivel 1 — Tendencias globales de mortalidad, 2000–2024"
date: today
lang: en
---

<span class="badge-nivel">Bloque A · Nivel 1</span>
**Variables** : `year` · `sex`

## Introducción

Este nivel constituye la **puerta de entrada** al análisis. Se examinan las
tendencias brutas de mortalidad en México sin desagregacion (por causa ni territorio),
con el objetivo de identificar patrones macro y puntos de inflexión temporales.

---

```{r setup-a1, include=FALSE}
 load("H:/Mon Drive/Broni/Projet R Mortality/BDD/Data_clean/National/20260421 asmr/asmr_nac.rda")

library(dplyr)
mort <- asmr_nac |>   mutate(year = as.integer(year))

library(tidyverse); library(scales); library(knitr); library(kableExtra)

palette_sexo <- c("1" = "#2a7a8c", "2" = "#e84855", "Total" = "grey40")

# set.seed(42)
# mort <- expand.grid(
#   year  = 2000:2022,
#   sex   = c("Hombre", "Mujer"),
#   state = c("Ciudad de México", "Jalisco", "Nuevo León", "Veracruz",
#             "Chiapas", "Oaxaca", "Guanajuato", "Michoacán", "Puebla", "Sonora"),
#   cause = c("Enfermedades cardiovasculares", "Tumores malignos",
#             "Diabetes mellitus", "Enfermedades respiratorias", "Causas externas", "Otras")
# ) |> mutate(
#   deaths = rpois(n(), lambda = 500 + 40*(year-2000) + ifelse(sex=="Hombre",120,0)),
#   rate   = deaths / runif(n(), 1e5, 2e6) * 1e5
# )

mort_yr_sex <- mort |> group_by(year, sex) |> summarise(deaths=sum(n_death), .groups="drop")
mort_yr     <- mort |> group_by(year)      |> summarise(deaths=sum(n_death), .groups="drop") |>
               mutate(sex = "Total")
mort_global <- bind_rows(mort_yr_sex, mort_yr)
```

## Tendencias globales

::: {.panel-tabset}

#### Serie temporal

```{r fig-serie-temporal, fig.cap="Evolución de la mortalidad total (2000–2024)"}
ggplot(mort_global, aes(x=year, y=deaths/1000, color=sex, group=sex)) +
  geom_line(linewidth=1.1) + geom_point(size=2.2) +
  scale_color_manual(values=palette_sexo) +
  scale_x_continuous(breaks=seq(1990,2024,4)) +
  scale_y_continuous(labels=label_comma(suffix=" k")) +
 labs(title="Total Mortality in Mexico", subtitle="Deaths by Sex, 1990–2024",

x=NULL, y="Deaths (thousands)", color="Sex") +
  theme_minimal(base_size=13) +
  theme(plot.title=element_text(face="bold",color="#1a2744"), legend.position="bottom",
        panel.grid.minor=element_blank())
```

#### Tasa de crecimiento

```{r fig-crecimiento}
mort_yr |> mutate(growth = (deaths/lag(deaths)-1)*100) |> filter(!is.na(growth)) |>
  ggplot(aes(x=year, y=growth)) +
  geom_col(fill="#2a7a8c", alpha=0.8, width=0.7) +
  geom_hline(yintercept=0, color="#1a2744") +
  scale_x_continuous(breaks=seq(1990,2024,4)) +
 labs(title="Annual Mortality Growth Rate",
  x="Year", y="Annual Change (%)") +
  theme_minimal(base_size=13) +
  theme(plot.title=element_text(face="bold",color="#1a2744"), panel.grid.minor=element_blank())
```

:::

---


## ASMR trend

Statistical measure in epidemiology used to compare disease death rates across populations with different age structures.

ASMR Definition: Age-standardized mortality rate is used to compare death rates, particularly when comparing populations with different age structures.


::: {.panel-tabset}

#### Temporal serie
Trend Information: 
The results show consistent trends in age-standardized mortality from 1990 to 2024
Age-standardized mortality rates differ by sex, with males experiencing higher rates of disease-specific mortality compared to females.

for certain diseases

```{r fig-ASMR, fig.cap="ASMR trend (1990–2024)"}
ggplot(asmr_nac, aes(x=year, y=asmr_100k, color=sex, group=sex)) +
  geom_line(linewidth=1.1) + geom_point(size=2.2) +
  scale_color_manual(values=palette_sexo) +
  scale_x_continuous(breaks=seq(1990,2024,4)) +
  scale_y_continuous(labels=label_comma(suffix=" k")) +
 labs(title="ASMR trend in Mexico", subtitle="ASMR per 100k inhabitants by Sex, 1990–2024",

x=NULL, y="ASMR per 100k inhabitants", color="Sex") +
  theme_minimal(base_size=10) +
  theme(plot.title=element_text(face="bold",color="#1a2744"), legend.position="bottom",
        panel.grid.minor=element_blank())
```

#### Tasa de crecimiento

```{r fig}
# asmr_nac |> mutate(growth = (asmr/lag(asmr)-1)*100) |> filter(!is.na(growth)) |>
#   ggplot(aes(x=year, y=growth)) +
#   geom_col(fill="#2a7a8c", alpha=0.8, width=0.7) +
#   geom_hline(yintercept=0, color="#1a2744") +
#   scale_x_continuous(breaks=seq(1990,2024,4)) +
#  labs(title="Annual Mortality Growth Rate (asmr)",
#   x="Year", y="Annual Change (%)") +
#   theme_minimal(base_size=13) +
#   theme(plot.title=element_text(face="bold",color="#1a2744"), panel.grid.minor=element_blank())
```

---
## Navigation

::: {style="display:flex; justify-content:space-between; margin-top:2rem;"}
[← Inicio](index.qmd){.btn .btn-outline-secondary}
[→ 2- Desagregación por causa](A2_causas.qmd){.btn .btn-primary}
:::
