---
title: "⑤ Maximum disaggregation"
subtitle: "Block A · Level 5 — Health-territory articulation: state × specific cause × sex"
date: today
lang: en
sidebar: sidebar-en
---

<span class="badge-nivel">Block A · Level 5</span>
**Variables** : `year` · `sex` · `state` · `cause_specific`

## Introduction

This level represents the **maximum disaggregation** of Block A: it combines
all four analytical dimensions (time, sex, territory, and ICD-10 specific cause)
to produce high-resolution state–cause epidemiological profiles.

The central question is: **which state × cause × sex combinations concentrate
the greatest burden of preventable mortality?**

::: {.callout-tip}
**Recommended use**  
This level is the natural starting point for prioritising public health
interventions at the state level. Results are complemented by the spatial
analysis in [Block B](B2_spatial_states.qmd).
:::

---

```{r setup-a5-en, include=FALSE}
library(tidyverse); library(scales); library(ggrepel); library(kableExtra)
options(ggrepel.max.overlaps = Inf)

palette_sex <- c("1" = "#2a7a8c", "2" = "#e84855")

states_10 <- c("Mexico City", "Jalisco", "Nuevo León", "Veracruz",
               "Chiapas", "Oaxaca", "Guanajuato", "Michoacán", "Puebla", "Sonora")

cause_specific_list <- c(
  "Ischemic heart disease", "Heart failure", "Stroke",
  "Breast cancer", "Cervical cancer", "Prostate cancer", "Lung cancer",
  "Type 2 diabetes", "DM complications", "Liver cirrhosis",
  "Pneumonia", "COPD", "COVID-19",
  "Road traffic accidents", "Homicides", "Suicides"
)

set.seed(55)
mort5 <- expand.grid(
  year           = 1998:2024,
  sex            = c("1", "2"),
  state          = states_10,
  cause_specific = cause_specific_list,
  stringsAsFactors = FALSE
) |> mutate(
  deaths = rpois(n(), lambda = 80 + 8*(year-1998) +
                   ifelse(sex == "1", 30, 0) +
                   case_when(
                     cause_specific %in% c("Ischemic heart disease", "Type 2 diabetes") ~ 250,
                     cause_specific %in% c("Homicides", "Road traffic accidents")        ~ 150,
                     cause_specific == "COVID-19" ~ ifelse(year >= 2020, 300, 0),
                     TRUE ~ 40
                   ) +
                   case_when(
                     state == "Mexico City" ~ 200,
                     state == "Chiapas"     ~ -30,
                     TRUE ~ 0
                   )),
  rate = deaths / runif(n(), 5e5, 8e6) * 1e5
)

last_year <- max(mort5$year)
```

## State profiles by specific cause

::: {.panel-tabset}

#### Heatmap: state × specific cause

```{r fig-heatmap-a5-en}
mort5 |>
  filter(year == last_year) |>
  group_by(state, cause_specific) |>
  summarise(rate = mean(rate), .groups = "drop") |>
  mutate(cause_specific = str_wrap(cause_specific, 15),
         state          = str_trunc(state, 16)) |>
  ggplot(aes(x = cause_specific, y = reorder(state, rate), fill = rate)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1,
                       name = "Rate\n/100k") +
  labs(
    title    = paste0("Mortality rate by state and specific cause (", last_year, ")"),
    subtitle = "Age-standardised rate per 100 000 inhabitants",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x    = element_text(angle = 40, hjust = 1, size = 8),
        plot.title      = element_text(face = "bold", color = "#1a2744"),
        plot.subtitle   = element_text(color = "#6c757d"),
        panel.grid      = element_blank())
```

#### State ranking by cause

```{r fig-ranking-estatal-en}
top_cause <- mort5 |>
  filter(year == last_year) |>
  group_by(cause_specific) |>
  summarise(total = sum(deaths)) |>
  slice_max(total, n = 6) |>
  pull(cause_specific)

mort5 |>
  filter(year == last_year, cause_specific %in% top_cause) |>
  group_by(state, cause_specific) |>
  summarise(rate = mean(rate), .groups = "drop") |>
  mutate(cause_specific = str_wrap(cause_specific, 18)) |>
  ggplot(aes(x = rate, y = reorder(state, rate), fill = rate)) +
  geom_col(alpha = 0.85) +
  facet_wrap(~cause_specific, scales = "free_x", ncol = 3) +
  scale_fill_distiller(palette = "Blues", direction = 1, guide = "none") +
  labs(
    title = "State ranking for the 6 leading specific causes",
    x = "Rate / 100 000 inh.", y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(plot.title       = element_text(face = "bold", color = "#1a2744"),
        strip.text       = element_text(face = "bold", size = 8),
        panel.grid.minor = element_blank())
```

#### Gender gap by state and cause

```{r fig-gender-gap-en}
mort5 |>
  filter(year == last_year, cause_specific %in% top_cause[1:4]) |>
  group_by(state, sex, cause_specific) |>
  summarise(rate = mean(rate), .groups = "drop") |>
  pivot_wider(names_from = sex, values_from = rate) |>
  mutate(ratio_mf = `1` / `2`,
         cause_specific = str_wrap(cause_specific, 18)) |>
  ggplot(aes(x = ratio_mf, y = reorder(state, ratio_mf),
             color = ratio_mf > 1)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  facet_wrap(~cause_specific, ncol = 2) +
  scale_color_manual(values = c("TRUE" = "#1a2744", "FALSE" = "#e84855"),
                     labels = c("TRUE" = "M > F", "FALSE" = "F > M")) +
  labs(
    title  = "Male/female mortality rate ratio by state and cause",
    x = "M/F ratio (> 1 = male excess)", y = NULL, color = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(plot.title       = element_text(face = "bold", color = "#1a2744"),
        strip.text       = element_text(face = "bold"),
        legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

:::

---

## Temporal trends by state and cause

```{r fig-temporal-trends-en, fig.height=9}
states_sel <- c("Mexico City", "Jalisco", "Chiapas", "Nuevo León")

mort5 |>
  filter(state %in% states_sel, cause_specific %in% top_cause[1:3]) |>
  group_by(year, state, sex, cause_specific) |>
  summarise(rate = mean(rate), .groups = "drop") |>
  mutate(cause_specific = str_wrap(cause_specific, 14)) |>
  ggplot(aes(x = year, y = rate, color = sex, group = sex)) +
  geom_line(linewidth = 0.9) +
  facet_grid(state ~ cause_specific, scales = "free_y") +
  scale_color_manual(values = palette_sex,
                     labels = c("1" = "Men", "2" = "Women")) +
  scale_x_continuous(breaks = c(2000, 2010, 2020)) +
  labs(
    title    = "Temporal trends: state × specific cause × sex",
    subtitle = "Selection of 4 states and 3 leading causes",
    x = NULL, y = "Rate / 100 000 inh.", color = "Sex"
  ) +
  theme_minimal(base_size = 10) +
  theme(plot.title       = element_text(face = "bold", color = "#1a2744"),
        plot.subtitle    = element_text(color = "#6c757d"),
        strip.text       = element_text(face = "bold", size = 8),
        legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

---

## Summary table

```{r tabla-sintesis-a5-en}
mort5 |>
  filter(year == last_year) |>
  group_by(state, cause_specific) |>
  summarise(
    total_deaths = sum(deaths),
    mean_rate    = round(mean(rate), 1),
    .groups = "drop"
  ) |>
  slice_max(total_deaths, n = 15) |>
  arrange(desc(total_deaths)) |>
  mutate(total_deaths = format(total_deaths, big.mark = ",")) |>
  knitr::kable(
    caption   = paste0("Top 15 state × cause combinations by mortality burden (", last_year, ")"),
    col.names = c("State", "Specific cause", "Deaths", "Rate /100k")
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"), full_width = FALSE
  ) |>
  kableExtra::row_spec(1:3, bold = TRUE, background = "#f0f4f8")
```

---

## Key findings

::: {.callout-note icon="false"}
**Key takeaways from Level ⑤**

- The state × cause × sex combination reveals **specific geographic concentrations**
  that remain hidden at more aggregated levels of analysis.
- **Ischemic heart disease** and **type 2 diabetes** dominate in almost every state,
  but with very different intensities depending on the region.
- **External causes** (homicides, accidents) show the highest M/F ratio and are
  concentrated in northern and north-western states.
- This level provides the entry point for **prioritising public health interventions**
  with maximum territorial specificity.
:::

---

## Navigation

::: {style="display:flex; justify-content:space-between; margin-top:2rem;"}
[← ④ Territorial dimension](A4_territorial.qmd){.btn .btn-outline-secondary}
[→ Block B · ① Regional description](B1_regional.qmd){.btn .btn-primary}
:::
